Typed Clojure

### 2013-07-29 schemeworkshop13.tex
More semantics, start Java conversions rules
more judgements, metafunctions

\begin{abstract}
Typed Racket has proved to be a valuable starting point for
a gradual type system for Clojure.
Building a similar type system for a new language gives the
designer some flexibility to repurpose and extend features.
This paper gives an overview of Typed Clojure, concentrating
on the extensions and differences from Typed Racket. We also
show where Typed Racket's features were particularly useful
for type checking non-trivial Clojure idioms.
\end{abstract}

@@

\section{Higher-rank polymorphism}

@@

\section{Using negative filters}

@@

\section{Multimethods}

@@

\section{Extending occurrence typing}

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
                      \alt \true{} \alt \false{} \alt \nil{}
                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt {\doexp {\e{}} {\e{}}}
                      \alt {\fnexp {\x{}} {\t{}} {\e{}}}
                &\mbox{Expressions} \\ \\
  \s{}, \t{}    &::=& (\class{}\ \overrightarrow{\t{}})
                      \alt (\ma{\Value\ \v})
                      \alt (\cup\ \overrightarrow{\t{}})
                      \alt (\cap\ \overrightarrow{\t{}})
                      \alt (\Not\ \t{})
                      \alt (\All\ [\overrightarrow{\x{}}] \t{})
                      \alt (\TFn\ [\overrightarrow{\x{}}] \t{})
                &\mbox{Types}
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
\end{altgrammar}
$$
\caption{Syntax of Expressions, Types}
\end{figure*}

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x} {\tau}}}
{ \judgement {\propenv{}} {{\hastype {\x{}} {\t{}}}} {{\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}}
             {\emptyobject{}} {\topprop{}}}

@@

\infer [T-False]
{}
{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Nil]
{}
{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Do]
{ \judgement {\propenv{}} {{\hastype {\e1} {\t1}}} {{\filterset {\prop{1+}} {\prop2}}} {\object{1}} {{\prop{f1}}}
\\
  \judgement {\propenv{}, \prop{f1}} {{\hastype {\e2} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object{}} {\prop{f2}}
  }
{ \judgement {\propenv{}} {{\hastype {\doexp {\e1} {\e2}} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object} 
             {{\andprop {\prop{f1}} {\prop{f2}}}}}

\infer [T-If]

%\infer [T-App]

@@

\section{Java interoperability}

@@

\subsection{Arrays}

>>>>

\begin{abstract}
Typed Racket is a valuable starting point for
a gradual type system that targets Clojure.
Building a similar type system for a new language gives the
designer some flexibility to repurpose and extend features.
This paper gives an overview of Typed Clojure, concentrating
on the extensions and differences from Typed Racket. We also
show where Typed Racket's features were particularly useful
for type checking non-trivial Clojure idioms.
\end{abstract}

@@

\section{A Practical Type System for Clojure}

\subsection{Higher-rank polymorphism}

@@

\subsection{Using negative filters}

@@

\subsection{Multimethods}

@@

\subsection{Extending occurrence typing}

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
                      \alt \true{} \alt \false{} \alt \nil{}
                      \alt \class{}
                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt
                      {\doexp {\e{}} {\e{}}}
                      \\ &\alt& 
                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                &\mbox{Expressions} \\ \\
  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
                      \alt {\Value \v{}}
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt (\cap\ \overrightarrow{\t{}})
                      \alt (\Not\ \t{})
                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}
                                   {\prop{}}}
                      \\ &\alt&
                      {\ArrayTwo {\t{}} {\t{}}}
                &\mbox{Types} \\ \\
  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
                      \alt {\impprop {\prop{}} {\prop{}}}
                      \alt {\andprop {\prop{}} {\prop{}}}
                      \alt {\orprop {\prop{}} {\prop{}}}
                      \alt \topprop{}
                      \alt \botprop{}
                &\mbox{Propositions} \\
  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
                      \alt \emptyobject{}
                &\mbox{Objects} \\
  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
                &\mbox{Paths} \\
  \pesyntax{}   &::=& \classpe{}
                &\mbox{Path Elements} \\ \\
                % TODO objects with path
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
  \Gamma       &::=& \overrightarrow{\prop{}}
               &\mbox{Environments}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions, and Objects}
\end{figure*}

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
             {\emptyobject{}} {\topprop{}}}

@@

\infer [T-False]
{}
{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Nil]
{}
{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Do]
{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
             {\flowprop {\prop{1}}}
\\\\
  \judgement {\propenv{}, {\flowprop {\prop{1}}}} {\hastype {\e2} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} 
             {\object{}} {\flowprop {\prop{2}}}
  }
{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e2}} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} {\object} 
             {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}}

\infer [T-If]
{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
             {\object{1}} {\flowprop {\prop{1}}} 
             \\\\
  \judgement {\propenv{}} {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
             {\object{}} {\flowprop {\prop{2}}}
  \\\\
  \judgement {\propenv{}} {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
             {\object{}} {\flowprop {\prop{3}}}
             }
{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
            % do we need to include prop1+/- here?
             {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
                         {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
             {\object{}} {\orprop {\flowprop {\prop{2}}}
                                  {\flowprop {\prop{3}}}} }

\infer [T-Let]
{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
             {\object{1}} {\flowprop {\prop{1}}}
  \\\\
  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
             {\object{}} {\flowprop {\prop{2}}}
             }
{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
                          {\object{1}}
                          {\x{}}}
             {\replacefor {\object{}} 
                          {\object{1}}
                          {\x{}}}
             {\replacefor {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}
                          {\object{1}}
                          {\x{}}}}

\infer [T-Throw]
{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
             {\object{}} {\flowprop {\prop{}}}}
{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
             {\emptyobject{}} {\botprop} }

\infer [T-Fn]
{}
{ TODO }

\infer [T-New]
{ }
{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
             {\filterset {\topprop{}} {\topprop{}}}
             {\emptyobject}
             {\topprop{}}}

@@

\begin{figure*}
$$
\begin{altgrammar}
{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
\\
{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
            {\javatotc {\class{2}} {\true{}}}
            {\filterset {\topprop{}} {\topprop{}}}
            {\emptyobject{}}
            {\topprop{}}}
\\
{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
            {\javatotc {\class{2}} {\false{}}}
            {\filterset {\topprop{}} {\topprop{}}}
            {\emptyobject{}}
            {\topprop{}}}
\\\\

{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
\\
{\javatotc {\class{}} {\false{}}} &=& \class{}
\\
{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \true{}}}
\\
{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \false{}}

\end{altgrammar}
$$
\caption{Converting Java types to Typed Clojure types}
\end{figure*}

\subsection{Java interoperability}

@@

\subsubsection{Arrays}
